The impact of gas bulk rotation on the Lyman-$\alpha$ line

### 2014-02-03 paper/ms.tex
fianlly FWHM plot included
my version

\begin{figure}
    \includegraphics[width=0.45\textwidth]{NewWidthVvsVmax.eps}
    \caption{Half-width for the non-dusty models as a function of
      rotational velocity $V_{\rm max}$. Continuous (dashed) lines
      correspond to central (homogeneous) source
      distributions. \label{fig:widthvsvelocity}} 
\end{figure}

@@

Rotation breaks the spherical symmetry of the static model and
introduces the rotation axis as a preferred direction. This motivates
us to study the anisotropy of the outgoing photons. In order to study
possible deviations with respect to an isotropic flux we follow
\cite{Zheng2013} to estimate the flux as a function of the angle
$\Theta$ formed by the outgoing photons and the observer. Figure
\ref{fig:OCspheres} presents a geometric scheme to clarify the
definition of all relevant directions in the problem.

@@

The off-centered cases with hints of triple peaked emision motivate us
to explore this in more detail. The first condition is to increase the
number of photons in the radiative transfer calculation. In the line
profiles presented in the previous subsection, only $\sim 10^4$ photons
have the geometric conditions to build the line profile. We run a few
models with $N_{\gamma} = 10^6$ in order to improve the statistics and
explore smaller radii for the off-centered spheres.

@@

Figure \ref{fig:3p_profiles} summarizes the results for all these
models. In all cases the stronger triple peaked signatures is
presented by the emitting spheres close centered at $0.75R$ close to
the surface. Emitting spheres closer to the center do not show clear
signatures for a triple peaked emission.

Furthermore, taking into account all the models that do not event present hints for
a triple peaked emission it would seem that the rotational velocity
and optical depth must balance each other in order to show this feature
feature. Namely, if the rotational velocity is close to half the HFWM
of the line, the triple peaks are more likely to occurr. This is the
same threshold that in the case of centered emission marks the
difference between double and single peaked lines.

@@

{\bf Figure \ref{fig:3p_res} show that even with a low resolution a three-peak profile 
would be observed, this is important for observational porpoises due to the finite 
spectral resolution. Seguro? Que es baja resolucion y que es alta
resolucion en este contexto?}

\begin{figure}
\begin{center}
\includegraphics[scale=0.3]{f12.eps}
\end{center}
\caption{Three-Peak profiles for different resolutions:Low (Solid), medium(dashed), high(doted), 
   \label{fig:3p_res}} 
\end{figure}

@@

We follow a classification of \ly profiles made by
\cite{Kulas12}. They presented observational results for star forming
galaxies at $z=2$-$3$ with multiple peaks in their \ly emission.
They classified the observed lines into five groups. In Group I the
red peak is stronger than the blue, while in Group II the blue peak is
stronger, in both cases the peaks are symmetrically located around the
line center. In Group III the blue peak also dominates but there is an
overall redward shift of the zero point. Group IV presents two similar
peaks symmetrically located around the zero point and Group V galaxies show
three peaks.

The asymmetries in Groups I, II  and III cannot be explained by
rotation. However, some galaxies in those groups have a high flux between
the two peaks; an effect that can be induced by rotation. The same
argument applies to galaxies in the group IV in which the flux
in between of the two peaks is high ($ \geq 70 \%$ of the intensity in
the highest peak). Finally, the triple peaked lines in Galaxies in
Group V can be easily reproduced by rotation with off-centered
emission given the right amount of asymmetry between the emitting
region and neutral hydrogen distribution.

Another relevant sample of observations was presented by \cite{Yamada2012} 
for 91 LAEs at $z=3.1$. They find that $\sim 33 \%$ of the galaxies present 
a single peak profile. Some of the peaks have a strong symmetry, while
others do not. The symmetric single peaked profile can be explained by
rotation effects, while the asymmetric cases naturally arise in
outflow/inflow models \citep{Verhamme2008, Dijkstra06}. Among the 91
LAEs, 8 reveal triple peaked profiles with a central peak and two
smaller peaks at the redder and bluer side, a feature that is present
in our off-center emission models with rotation Figure \ref{fig:3p_profiles}.

The presence of single peaked profiles has been associated to
inflow/outflow dynamics \citep{XXX}. In this paper we show that gas
bulk rotation can also be considered as a probable origin for that
behavior, provided that the observed single peak is highly
symmetric. Similarly, in the case of double peaked lines with a high
level of flux at the line center, rotation also deserves to be considered in
the pool of possible bulk flows responsible for that feature,
specially if the two peaks have similar intensities. The case of
triple peaked lines is another clear feature of gas rotation under the
additional condition of an off-centered emission.

@@

This paper quantifies for the first time in the literature the effects
of rotation in the morphology of the \ly emission line in star forming
galaxies.  The results are based on the study of an homogeneous sphere
of gas with solid body rotation. We explore a range of models by varying
the rotational speed, hydrogen optical depth, dust optical depth and
initial distribution of \ly photons with respect to the gas
density. As a cross-validation, we obtain our results from two
independently developed Monte-Carlo radiative transfer codes.

Our main result is that rotation clearly impacts the \ly line
morphology. Double peaked lines can make transitions into single
peaked lines as the rotational velocity increases. In the case of
off-centered emission result into triple peaked lines for certain
values of optical depth and rotational velocities.

@@

\item The line width increases with rotational velocity. The line with
  approximately follows the functional form $W^2 = W_{
    0}^2 + (V_{\rm max}/\lambda)^2$ where $W_{0}$ indicates the line
  width for the static case. $\lambda$ is a constat determined from
  the radiative transfer results it is $\lambda_{\rm c}=XX$ and
  $\lambda_{\rm h}=YY$ for the central and homogeneous source
  distribution, respectively.

\item A single peaked line emerges for high rotational velocities in
  the case of homgeneously distributed sources. The cases of single
  peaked lines occur when the rotational velocity is larger than the
  line width measured in the static case.

\item For central sources we find that the number of scatterings
does not significantly decrease with rotation, this is due to the 
fact that the vast majority of scatterings events are resonant, 
during which the mean free path is very short and the effect of gas
bulk rotation is not enough to affect \ly photons.
The escape fraction does not depend on the rotational velocity neither
the position of the peaks, but there is an increase 
in the flux on the central parts of the line for high rotational 
velocities.

\item For homogeneous sources we find that the changes in the
line morphology are linked to the reduction in the average number of
scatterings. In the second part of \S \ref{sec:widthpeak} we show
that, as rotational velocity increases, a large fraction of photons
escape with less than $10$ scatterings and on average 
with $40\%$ less scatterings than in the static case. This makes highly improbable 
that a photon's frequency can move away from the line's center. 
In \S \ref{sec:widthpeak} For typical column densities of $N_{HI}=1e^{19}-1e^{20}$, 
we find that rotational velocities of 100-300 \kms broaden the line profile up 
to a factor of $2-3$ in comparison to the static case. Simultaneously, the
flux in the line center also increases with rotational velocity. This
increase induces the transition from a double peak to a single peak
line in the models where the radiation sources are homogeneously
distributed. Also the escape fraction increase by factors of $2-10$ with 
respect to the slab with centrally sources.

\item For non-symmetric sources we find that three peak profiles
could emerge for models in which the rotational velocity is the necessary 
to give rise to a central peak, but not that high to decrease the
amount of photons in the wings. When initial off-center sources are 
selected from the outer parts of the galaxy the wings of the line are stronger.
Also we do not find any considerable asymmetries in the line produced by off-center 
emission.
\end{itemize}

Comparing our results with recent observed LAEs we find that many 
morphological features such as high central line flux, single peak
profiles and multi-peaked profiles could be explained by gas bulk 
rotation present in these LAEs.

>>>>

\begin{figure}
    \includegraphics[width=0.45\textwidth]{f2.eps}
    \caption{Half-width for the non-dusty models as a function of
      rotational velocity $V_{\rm max}$. Continuous (dashed) lines
      correspond to central (homogeneous) source
      distributions. \label{fig:widthvsvelocity}} 
\end{figure}

@@

Rotation breaks the spherical symmetry of the static model and
introduces the rotation axis as a preferred direction. This motivates
us to study deviations from an isotropic flux for the outgoing
photons. We follow \cite{Zheng2013} to estimate the flux as a function
of the angle $\Theta$ formed by the outgoing photons and the observer. Figure
\ref{fig:OCspheres} shows a geometric scheme with all the angles and
vectors relevant to this problem.

@@

We now study in more detail the conditions for the triple peaked
emission. To do this we perform new calculations with $N_{\gamma} =
10^6$ in order to improve the statistics and explore smaller radii for
the off-centered spheres.

@@

Figure \ref{fig:3p_profiles} summarizes the results for all these
models. In all cases the stronger triple peaked signatures are
present in the emitting spheres close to the surface, centered at
$0.75R$. Emitting spheres closer to the center do not show
clear signatures for a triple peaked emission.

Taking into account all the models that do not show any hint for
a triple peaked emission suggests that the rotational velocity
and optical depth must balance each other in order to show this feature
feature. If the rotational velocity is close to half the HFWM
of the line in the static case, the triple peaks are more likely to
occurr. This is close to the threshold that marks the difference
between double and single peaked lines.

@@

%{\bf Figure \ref{fig:3p_res} show that even with a low resolution a
%  three-peak profile  would be observed, this is important for
%  observational porpoises due to the finite  spectral
%  resolution. Seguro? Que es baja resolucion y que es alta resolucion en este contexto?}

%\begin{figure}
%\begin{center}
%\includegraphics[scale=0.3]{f12.eps}
%\end{center}
%\caption{Three-Peak profiles for different resolutions:Low (Solid), medium(dashed), high(doted), 
%   \label{fig:3p_res}} 
%\end{figure}

@@

We follow a classification of \ly profiles made by
\cite{Kulas12}. They presented observational results for $18$ star forming
galaxies at $z=2$-$3$ with multiple peaks in their \ly emission.
They classified the observed lines into five groups. In Group I the
red peak is stronger than the blue, while in Group II the blue peak is
stronger, in both cases the peaks are symmetrically located around the
line center. In Group III the blue peakdominates but there is an
overall redward shift of the velocity zero point. Group IV presents two similar
peaks symmetrically located around the zero point and Group V galaxies show
three peaks.

The asymmetries in Groups I, II  and III cannot be explained by
rotation. However, some galaxies in those groups have a high flux between
the two peaks; an effect that can be induced by rotation. The same
argument applies to galaxies in the group IV in which the flux
in between of the two peaks is high ($ \geq 70 \%$ of the intensity in
the srongest peak). Finally, the spectrum of the single galaxy in
Group V out of the total $18$ can be produced by rotation with
off-centered emission given the right amount of asymmetry between the
emitting region and neutral hydrogen distribution.

Another relevant sample of observations was presented by \cite{Yamada2012} 
for $91$ LAEs at $z=3.1$. They find that $\sim 33 \%$ of the galaxies present 
a single peak profile. Some of the peaks have a strong symmetry, while
others do not. The symmetric single peaked profile can be explained by
rotation effects, while the asymmetric cases naturally arise in
outflow/inflow models \citep{Verhamme2008, Dijkstra06}. Among the 91
LAEs, 8 reveal triple peaked profiles with a central peak and two
smaller peaks at the redder and bluer side, a feature that is present
in our off-center emission models with rotation Figure \ref{fig:3p_profiles}.

The presence of single peaked profiles has been associated to
inflow/outflow dynamics \citep{XXX}. In this paper we show that gas
bulk rotation can also be considered as a probable origin for that
behavior, provided that the observed single peak is highly
symmetric. Similarly, in the case of double peaked lines with a high
level of flux at the line center, rotation also deserves to be considered in
the pool of possible bulk flows responsible for that feature,
specially if the two peaks have similar intensities. The case of
triple peaked lines is clear feature of gas rotation under the
additional condition of an off-centered emission.

@@

This paper quantifies for the first time in the literature the effects
of rotation in the morphology of the \ly emission line in star forming
galaxies.  The results are based on the study of an homogeneous
gaseous sphere with solid body rotation. We explore a range of models by varying
the rotational speed, hydrogen optical depth, dust optical depth and
initial distribution of \ly photons with respect to the gas
distribution. As a cross-validation, we obtain our results from two
independently developed Monte-Carlo radiative transfer codes.

Our main result is that rotation clearly impacts the \ly line
morphology. Double peaked lines can make transitions into single
peaked lines as the rotational velocity increases. In the case of
off-centered emission triple peaked lines are also possible for
certain values of optical depth and rotational velocities.

@@

\item The line width increases with rotational velocity, following the
  functional form $W^2 = W_{0}^2 + (V_{\rm max}/\lambda)^2$ where
  $W_{0}$ indicates the line width for the static case. $\lambda$ is a
  constant determined from the radiative transfer results it is
  $\lambda_{\rm c}=XX$ and $\lambda_{\rm h}=YY$ for the central and
  homogeneous source distribution, respectively.

\item For homogeneously distributed sources the double peaked line
  transforms into a single peaked line for high rotational
  velocities. This occurs when the rotational velocity is larger than half the
  line width measured in the static case. In the case of central
  sources the double peaked line was always present in the range of
  explored rotational velocities.

\item For homogeneously distributes sources the average number of
  scatterings before escaping decreases with rotational velocity, this
  is driven by the existence of a photo-sphere where photons can
  escape after a single scattering. For central sources this
  photo-sphere does not exist and we do not find any significant
  change in the average number of scatterings.

\item As a consequence of the change in the average number of
  scatterings there is an impact on the escape fraction. Consistent
  with the picture summarized in the previous item, the escape
  fraction increases (up to a factor of $XXX$) with rotational
  velocity for the spheres with homogeneous sources, while it stays
  constant in the case of central sources.

\item We explore the deviations in photon intensity from an isotropic
  flux. We find that the variations are always less than $5\%$ with
  respect to the homogeneous emission from the

\item In the case of off-centered emission it is possible to find
  spectra with a triple peaked profile. This feature is clearly present for
  sources emitted close to the surface of the sphere in cases where
  the rotational velocity is half the FWHM measured in the static case.

\item Comparing our results with recent detailed observations of the
  \ly profile at redshifts $z\sim 2-3$ we find that morphological
  features such as high central line flux, single peak profiles and
  multi-peaked profiles could be explained by gas bulk  rotation
  present in these LAEs. 
\end{itemize}
